Depolarization of Interacting NV ensembles : \\ The Role of Local Electric Fields and Double Flip processes

### 2022-06-21 Mes papiers/Low field CR/Low_field_CR.tex
tutu

\title{Low magnetic field depolarization of NV$^-$ centers through dipole-dipole interaction}

@@

\begin{abstract}
C'est trop bien
\end{abstract}

@@

Intro

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig dense vs pas dense}
\caption{Photoluminescence measurement from NV centers ensemble as a function of an external magnetic field randomly oriented (a) for a CVD sample containing $\approx$ 4 ppb NV$^-$ centers, (b) for an HPHT sample containing $\approx$ 3 ppm NV$^-$ centers}
\label{PL_NV_density}
\end{figure}

@@

Fig. \ref{PL_NV_density} shows the change in photoluminescence (PL) with respect to an external magnetic field for two samples with a low and high concentration of NV$^-$ centers. Both samples show a decrease in PL as the magnetic field amplitude increases, due to eigenstate mixing caused by the transverse part of the magnetic field \citep{epstein2005anisotropic,lai2009influence}. We notice however a stark difference in the low magnetic field region where only the high-density sample observe a drop in PL. In this article we propose to explore the reasons behind this zero-field line present only for dense NV$^-$ ensemble.

Unlike the decrease in PL for higher magnetic field values, the drop in PL at low magnetic field is associated with a decrease of the NV's spin lifetime $T_1$. Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}. To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig T1}
\caption{Spin $T_1$ measurement protocol. (a) Sequence used to measure the spin lifetime : the laser is turned off for a time $\tau$ twice and the spin state is read at the end of the dark time by measuring the initial photoluminescence when the laser is turned back on. A microwave $\pi$-pulse resonant with one of the spin transition is placed before the second reading. The  $T_1$ sequence result is the subtraction $S_1-S_2$. (b) Measurement of $S_1-S_2$ for a high density of NV centers in zero magnetic field field and non-zero magnetic field. The fit (orange plain line) is detailed in the main text}
\label{T1}
\end{figure}

The $T_1$ depolarization dynamics of single or dilute NV centers at room temperature is dominated by a two-phonon Raman process \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depends on the crystal lattice temperature but does not depend on the external magnetic field. However, it has been observed that dense NV centers ensemble have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi_depolarization_2017, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field and not on the temperature . This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi_depolarization_2017}. Some inhomogeneity between the NV centers is further needed in order to explain the depolarization of the spin ensemble. We will denote $T_1^{\rm ph}$ the characteristic time associated with the phonon relaxation process and $T_1^{\rm dd}$ the characteristic time associated with the dipole-dipole relaxation process.

The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi_depolarization_2017}. It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. Using only this sequence can result in artifacts in the signal, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right before the spin read-out to project the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers.

\begin{figure*}
\includegraphics[width=.95\textwidth]{Figures/fig 100 vs 1x1x1x1}
\caption{Dependency on the magnetic field angle for the zero field depolarization. (a) ODMR spectrum in zero field. (bi) ODMR spectrum for a magnetic field $\approx$ 60 G, misaligned by $\sim  24^\circ$ with the [100] axis. (bii) Normalized photoluminescence of the NV$^-$ ensemble as a function of the magnetic field amplitude. (biii) Stretch part of the spin decay $1/T_1^{\rm dd}$ as a function of the magnetic field amplitude. (ci), (cii) and (ciii) : same measurement with a magnetic field aligned with the [100] axis}
\label{100_VS_1x4}
\end{figure*}

The ensemble spin decay is non-exponential due to inhomogeneities between the NV$^-$ centers. We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi_depolarization_2017}. A conclusion of this model is that, for an homogeneous 3D distribution of fluctuator, the dipole-induced lifetime should be stretched-exponential with a stretch factor $\beta=1/2$. Such an ensemble lifetime is indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$  (see SI).

For the samples used in this paper (detailed in SI), we observed that $T_1^{\rm dd} \sim T_1^{\rm ph}$, meaning that we had to include both  in our analysis. The resulting fitting formula is therefore :
\begin{equation}
S(\tau)=A \exp (-\frac{\tau}{T_1^{\rm ph}} -\sqrt{\frac{\tau}{T_1^{\rm dd}}}),
\end{equation}
Further details on the fitting procedure are available in SI. For the rest of the article we will be only interested on the stretched exponential lifetime ${T_1^{\rm dd}}$.

%\begin{figure}
%\includegraphics[width=0.45\textwidth]{Figures/fig largeur fluct}
%\caption{Tuning resonant coupling through magnetic field orientation. (a) Sketch of the four possible orientations ("classes") in a single crystal diamond lattice. (b) ODMR spectrum showing four $\ket{0} \to \ket{-1}$ resonances corresponding to the four spin classes. The detuning $\Delta \nu$ between the classes $\beta$ and $\gamma$ was controlled by changing the orientation of the external magnetic field.(c) Stretch part of the lifetime decay for the spins resonant with $\nu_\gamma$ as a function of the detuning $\Delta \nu$ (blue circles), fitted by a Lorentzian with half width at half maximum 8.04 MHz . Green dashed line correspond to the ODMR width of a single class stretched by a factor $\sqrt 2$, approximating the overlap of the two classes (see SI).}
%\label{largeur_fluct}
%\end{figure}

%Fig. \ref{largeur_fluct} illustrates how we can use the NV center geometry to tune the resonant dipolar-coupling between the spins. Because of the diamond symmetry, we can distinguish 4 groups ("classes") of NV centers, based on the four possible orientation of the NV axis in the crystal. These four classes have \textit{a priori} different spin resonance frequency based on the projection of the magnetic field on their respective NV axis. By tuning the angle of the magnetic field, we can therefore control the frequency detuning between the different classes, and as a result we can control the number of resonant NV centers by bringing in and out of resonance several classes.
%
%Fig. \ref{largeur_fluct} (c) in particular shows the dipole-induced lifetime $T_1^{\rm dd}$ for a given class as a function of the energy detuning with another class. One can notice that the width of the $T_1^{\rm dd}$ bell shape is significantly larger than that of the frequency overlap between the two classes, modeled here by an ODMR line stretched by a factor $\sqrt{2}$ (see SI). This extra width can not be attributed to the dipole-dipole coupling strength : for a 3 ppm concentration of NV$^-$ centers, the average coupling strength between neighboring NV centers should be $\sim 27\ \rm kHz$, two order of magnitude below the observed width. We can also see that $T_1^{\rm dd}$ follows a Lorentzian shape despite the ODMR lines being non-Lorentzian. These two observations corroborate the idea developed in \cite{choi_depolarization_2017} that the interaction width is dominated by the noise spectrum of the lifetime-limited fluctuators.

One of the way we can isolate dipole-dipole related phenomena with NV centers is by tuning the number of NV centers resonant with each other. Indeed, Fig. \ref{100_VS_1x4} (bi) shows the characteristic 8 lines observed with ensemble NV optically detected magnetic resonance (ODMR) spectrum. These 8 lines correspond to the four "classes" of NV centers - that is physical orientation of the NV axis is the diamond crystal cell - multiplied by the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions of the four different classes can be controlled somewhat independently by changing the amplitude and orientation of the external magnetic field, and in particular they can be brought to resonance for particular orientation of the magnetic field. Fig. \ref{100_VS_1x4} (ci) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. Other example of class resonance are given in SI.

Fig. \ref{100_VS_1x4} (bii) and (biii) shows respectively the PL and the stretched exponential spin decay $T_1^{\rm dd}$ of an ensemble of NV centers with respect to the amplitude of the external magnetic field. The orientation of the magnetic field was chosen in order to lift the degeneracy between the four classes, as shown in Fig. \ref{100_VS_1x4} (bi). We observe that the drop in PL at low magnetic field is indeed correlated with an increase in the spin decay rate, as is expected due to the higher number of resonant spins in zero magnetic field, where all four classes are resonant, compared to the non-degenerate case for $\vec \neq 0$. We can also see that the decrease in PL for $B>40\ \rm G$ due to state mixing by the transverse magnetic field is not correlated to a modification of the spin lifetime.

However, the splitting of the four classes is not sole reason to the decrease of the spin lifetime in low magnetic field. Indeed, we performed similar measurements in Fig. \ref{100_VS_1x4} (cii) and (ciii) but this time fixing the magnetic field orientation along the [100] axis. For this particular orientation, the four classes are always resonant, regardless of the field amplitude. We can see that, although considerably reduced, there still is an increase in the spin decay rate and a corresponding drop in PL for low magnetic field values. We attribute the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20\ \rm G$ to dipolar interaction with NV having a first neighbor $^{13}C$ \cite{pellet2021optical}.

The aim of this paper is to understand the remaining contribution to the spin decay in this second case. We identified and isolated two additional effects : the domination by the local electric field for low magnetic field value which results in a change of the spin Hamiltonian eigenstates, and the presence of double-flip processes where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.

In order to distinguish these two contributions, we decided to simulate the effect of the electric field by using a transverse magnetic field : for relatively weak transverse magnetic field, the eigen basis of the spin Hamiltonian is close to $\{ \ket{0},\ket{+}=\frac{\ket{+1}+\ket{-1}}{\sqrt{2}},\ket{-}=\frac{\ket{+1}-\ket{-1}}{\sqrt{2}} \} $, which is the the spin Hamiltonian basis when the electric field dominates, as detailed in SI. This property has been used to observe a linear dependence in the electric field for the spin transition frequencies by applying an external transverse magnetic field \cite{dolde2011electric,qiu2022nanoscale}.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig transverse field V2}
\caption{(a) Simulated energies for the spin Hamiltonian in the presence of pure transverse magnetic field $B_\perp$. The three spin levels are denoted $\ket{g}$, $\ket{d}$ and $\ket{e}$. (b) Blue dashed curve : frequency detuning $\Delta \nu$ between the $\ket{g}$, $\ket{d}$ states. Red plain curve : Matching factor $\abs{\bra{e}\ket{+}}^2$ between the states $\bra{e}$ and $\bra{+}$. (c) Measurement of the stretched exponential decay rate $1/T_1^{\rm dd}$ for a single class as a function of the measured detuning $\Delta \nu$ between the two transitions $\ket{g} \to \ket {d}$ and  $\ket{g} \to \ket {e}$ for purely transverse magnetic field. The green dashed line correspond to the decay rate for the same class when the magnetic field is longitudinal}
\label{B_transverse}
\end{figure}

Fig. \ref{B_transverse} (a) shows the energy levels for the three eigenstates of the NV spin Hamiltonian with respect to a purely transverse magnetic field. We will denote these eigenstates in the general case $\ket{g}$, $\ket{d}$ and $\ket{e}$. For weak transverse magnetic field, $\ket{g} \approx \ket{0}$, $\ket{d}=\ket{-}$ and $\ket{e} \approx \ket{+}$. As the magnetic field increases, $\ket{g}$ and $\ket{e}$ starts to become a mixing of $\ket{0}$ and $\ket{+}$ while $\ket{d}$ remains equal to $\ket{-}$.

Crucially, this mixing happens on a different scale than the associated energy splitting of the $\ket{d}$ and $\ket{e}$ states : Fig. \ref{B_transverse} (b) shows both the splitting between the energy levels of $\ket{d}$ and $\ket{e}$, and the matching factor $\abs{\bra{e}\ket{+}}^2$ characterizing the closeness of the states $\bra{e}$ and $\bra{+}$. We can see that for a field value of $\sim 130 G$, the splitting between the $\ket{d}$ and $\ket{e}$ reaches $\sim 50 MHz$, far exceeding the range of the dipole-dipole resonant coupling (see details in SI), but $\abs{\bra{e}\ket{+}}^2 > 0.98$ meaning that the spin Hamiltonian eigenstates are still roughly in the $\ket{0}, \ket{+}, \ket{-}$ basis. We can therefore decouple the effects of the double-flip processes, which are no longer resonant, from the effects of the change of basis induced by the transverse electric - or in this case magnetic - field.

Fig. \ref{B_transverse} (c) shows the measurement of the stretched lifetime $T_1^{\rm dd}$ for a single class of NV centers exposed to a transverse magnetic field between 25 and 130 G. The detuning $\Delta \nu$ between the two states $\ket{d}$ and $\ket{e}$ was measured through ODMR. We attribute the decrease in the spin decay rate when the detuning increases to the loss of effectiveness of the double-flip processes. We can notice that the decay rate reaches a plateau for $\Delta \nu \geq 30\ \rm MHz$ with a value about twice as big as the decay rate for a longitudinal magnetic field. This plateau we attribute to the effect of the transverse magnetic field, which we assume to be similar to the effect of the electric field in the low magnetic field regime. We should note that the effect of the double-flip processes is about 4 times more important in the depolarization rate than the effect of the electric field, even though the states are not initially fully resonant.

Overall we predicted, based on an extension of the fluctuator model developed in SI, an increase by a factor of $\sim 4$ of the dipole-dipole induced decay rate in the non-magnetic basis $\ket{0}, \ket{+}, \ket{-}$ compared to the magnetic basis $\ket{0}, \ket{+1}, \ket{-1}$. While the measured increase is smaller than the predicted one, as with most of the predictions based on the fluctuator model, the measurements agrees with the theoretical increase in the flip-flop rate in the $\ket{+/-}$ basis compared to the $\ket{\pm 1}$ basis.

Similarly, we can compute the predicted decay rate in the $\vec B \parallel \left[100\right]$ case compared to the $\vec B = 0$ case, and find a theoretical increase of $\sim 20\%$ in the $\vec B = 0$ case due to the effects of the local electric field. This, along with the double-flip processes explains the behavior observed in Fig. \ref{100_VS_1x4} (cii) and (ciii).

>>>>

\title{Depolarization of Interacting NV ensembles : \\
The Role of Local Electric Fields and Double Flip processes}

@@

\begin{abstract}
We present experimental results on dipolar interaction processes in ensemble of spin-1 systems. Using the spin of highly doped ensembles of negatively charged nitrogen vacancy centers in diamonds, we identify regimes where flip-flop, double-flip processes as well as the mixing induced by local electric field dominate.
%We analyse these results from the perspective of magnetometry applications, and show that both double-flip processes and local electric fields improve the magnetic field sensitivity.
Our results are relevant for understanding decoherence in many-body spin systems as well as for high sensitivity magneto- and electro-metry with long-lived interacting solid-state spins. As a proof of principle, we present an orientation-free and microwave-free magnetometer operating with a sensitivity below $100$ nT/$\sqrt{\rm Hz}$.
\end{abstract}

@@

%\begin{figure}
%\includegraphics[width=0.45\textwidth]{Figures/fig_dense_vs_pas_dense.pdf}
%\caption{Photoluminescence measurement from NV centers ensemble as a function of an external magnetic field applied in an arbitrary direction (a) for a CVD sample containing $\approx$ 4 ppb NV$^-$ centers, (b) for an HPHT sample containing $\approx$ 3 ppm NV$^-$ centers}
%\label{PL_NV_density}
%\end{figure}

@@

The electronic spin properties of the negatively charged nitrogen-vacancy (NV$^-$) center in diamond has given rise to a wealth of applications in nanoscale sensing and quantum information science in part thanks to the possibility to optically polarize and read-out its spin state at ambient conditions. Many recent work indeed focus on ensembles of spins from NV centers with the prospect to boost the magnetic field sensing capabilities and for studying many body effects (resultats de chez BOSCH, LUKIN , FEFOR, ALEX, BIFONE)

When the spin ensemble becomes dense, inhomogeneities in the relaxation rates of NVs can give rise to spin depolarisation with a very rich many body dynamics associated with disorder. This mechanisms however limits the efficiency of magnetometers (LUKIN PRX).
%It was proposed to exploit these CR to perform sensitive microwave free magnetometry (REF).  
Recent magnetometry proposals have been put forward at zero magnetic field. There, the depolarisation effect was shown to be much stronger, but the detailed studies of relaxation mechanisms have not been identified to the best of our knowledge. Some of the possible processes that lead to cross-relaxation in strongly coupled dipolar systems are depicted in Fig.~1. The goal of this paper is to identify regimes where the spin flip-flop within different classes of NV centers, double-flip up and down as well as the mixing induced by local electric field play a role.

The electronic spin of NV center is a spin-1 system in the ground state (see Fig. 1). It can be optically polarized in the $\ket{m_s=0}$ state. The photoluminescence of this state is also is larger than the $\ket{m_s=\pm 1}$ states enabling spin-read out at ambient temperature. The $\ket{m_s=\pm 1}$ spin states are separated from the $\ket{m_s=0}$ state by  $D = (2\pi) 2.87$ GHz so that when a resonant microwave or static transverse magnetic field is applied, the photoluminescence is reduced \citep{epstein2005anisotropic,lai2009influence}. 
Fig.  \ref{T1}-a) and b) show the change in photoluminescence (PL) with respect to an external magnetic field for two samples with a low and high concentration of NV$^-$ centers (see SI). Both samples show a decrease in PL as the magnetic field amplitude increases.  There is however a stark difference in the low magnetic field region where only high-density samples shows a drop in PL (CITATION). This effect can be observed on all our samples (see SI) whenever the NV concentration lies in the ppm range.

The drop of the PL at low magnetic field is generally associated with a decrease of the NV's spin lifetime $T_1$ (CITE JARMOLA), which results in a decrease of the population in the bright state $0$.
%Indeed, since the $\ket{m_s=0}$ spin state is brighter than the $\ket{m_s=\pm 1}$ states, and since the optical pumping in the $\ket{m_s=0}$ state is in competition with the relaxation in a thermal distribution of the states, decreasing the spins lifetime results in a decrease of the NV PL \citep{finco2021imaging}.
%To understand the behavior of the PL in zero-field, we therefore need to study the dynamics of the spin in the low magnetic field region.
The depolarization dynamics of the spin of single or dilute NV centers at room temperature is dominated by two-phonon Raman processes \citep{redman1991spin,jarmola_temperature-_2012,norambuena2018spin}, which depend on the crystal lattice temperature. It has been observed that dense NV centers ensemble have an additional spin decay channel \citep{jarmola_temperature-_2012,jarmola_longitudinal_2015,mrozek_longitudinal_2015, choi_depolarization_2017, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, pellet2021magnetic, mrozek2021characterization}, which depends greatly on the magnetic field amplitude. This effect has been attributed to cross-relaxation between the NV centers through dipole-dipole coupling \citep{mrozek_longitudinal_2015, choi_depolarization_2017}.  Inhomogeneity of the spin lifetimes is further needed in order to explain the depolarization of the spin ensemble. We will denote $T_1^{\rm ph}$ the characteristic timescale associated with the phonon relaxation process and $T_1^{\rm dd}$ the density dependent timescale associated with the dipole-dipole relaxation process.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/shema_summary.pdf}
\caption{Schematics showing a diamond with interacting spins as well as the three processes that can account for dipolar relaxation. }
\label{T1}
\end{figure}

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig_T1.pdf}
\caption{a) Photoluminescence measurement from NV centers ensemble as a function of an external magnetic field applied in an arbitrary direction (a) for a CVD sample containing $\approx4$ ppb of NV$^-$ centers, (b) for an HPHT sample containing $\approx3$ ppm of NV$^-$ centers. 
(c) Sequence used to measure the spin lifetime. (d) Spin relaxation $S_1-S_2$ measured for the dense sample at zero and non-zero magnetic fields. The fitting procedure (orange plain line) is detailed in the main text}
\label{T1}
\end{figure}

\begin{figure*}
\includegraphics[width=.95\textwidth]{Figures/fig_100_vs_1x1x1x1.pdf}
\caption{(a) ODMR spectrum in zero field. (b)-i) ODMR spectrum for a magnetic field $\approx$ 60 G, misaligned by $\sim  24^\circ$ from the [100] axis. (b)-ii) Normalized photoluminescence of the NV$^-$ ensemble as a function of the magnetic field amplitude. (b)-iii) Stretched part of the spin decay $1/T_1^{\rm dd}$ as a function of the magnetic field amplitude. (c)-i), (c)-ii) and (c)-iii) : same measurements but with a magnetic field close to the [100] axis}
\label{100_VS_1x4}
\end{figure*}

Fig. \ref{T1}-c) shows the sequence employed for measuring $T_1^{\rm dd}$ in the dense sample. 
%The spin lifetime protocol used here is described in Fig. \ref{T1} and is based on previous similar experiments 
It consists in a pump-probe measurement where the spins are first polarized in the $\ket{0}$ state by a green laser, and read-out optically after a variable dark time $\tau$. In highly doped samples, this sequence often results in artifacts, mostly due to charge state transfer in the dark \citep{giri_coupled_2018, giri_selective_2019}. It is therefore convenient to repeat the sequence with an additional $\pi$ pulse right before the spin read-out to prepare the remaining $\ket{0}$ polarization into a darker $\ket{+1}$ or $\ket{-1}$ state. By subtracting the result of the two sequences, we select only the spin-dependent part of the signal, with the added benefit of being able to select a specific class of NV centers \citep{jarmola_temperature-_2012, mrozek_longitudinal_2015, choi_depolarization_2017}.

Fig. \ref{T1}-b) shows the subtracted signals when $\bm B=0$ and $\bm B \neq 0$.
The ensemble spin decay curves are not exponential, which was found to be due to inhomogeneities between the spin lifetimes of dipolar coupled NV$^-$ centers \citep{choi_depolarization_2017}.
We will base our interpretations of the experimental results on the NV-fluctuator model developed in \citep{choi_depolarization_2017}. A conclusion of this model is that, for a homogeneous 3D distribution of fluctuators, the dipole-induced lifetime should be stretched-exponential with a stretch factor $\beta=1/2$. Such an ensemble lifetime is indeed observed when $T_1^{\rm dd} \ll T_1^{\rm ph}$.
For the samples used in this paper (detailed in SI) $T_1^{\rm dd} \sim T_1^{\rm ph}$, so both decay processes have to included in the analysis. Fitting the two curves by the product of an exponential and a stretched exponential, we find $T_1^{\rm ph}=3.2$ ms and $T_1^{\rm dd}= ??$ here. We also conducted further measurements that confirm the lifetime-limited character of the fluctuator employed in the model \citep{choi_depolarization_2017}, see SI. 
%These results show that our sample clearly manifests dipolar interactions amongst closely packed NV centers, but do not tell about the magnitude of the several possible mechanisms involved in the dipolar relaxation. 
%In spin-1 systems, there can indeed be several channels for decay at zero fields such as flip-flop, double flip,

%The resulting fitting formula is therefore :
%\begin{equation}
%S(\tau)=A \exp (-\frac{\tau}{T_1^{\rm ph}} -\sqrt{\frac{\tau}{T_1^{\rm dd}}}),
%\end{equation}
%Further details on the fitting procedure are available in SI. For the rest of the article we will be only interested on the stretched exponential lifetime ${T_1^{\rm dd}}$.

One of the ways to tune dipole-dipole related phenomena is by changing the number of resonant spins. Fig. \ref{100_VS_1x4} (b)-i) shows an optically detected magnetic resonance (ODMR) spectrum from the dense ensemble. The observed lines correspond to the projections of the magnetic fields onto the four "classes" of NV centers - that is physical orientation of the NV axis is the diamond crystal cell - together with the two possible spin transitions $\ket{0}\to\ket{-1}$ and $\ket{0}\to\ket{+1}$. The transitions frequencies of the four different classes can be controlled by changing the amplitude and orientation of the external magnetic field. Fig. \ref{100_VS_1x4} (c)-i) shows an example where all four classes are brought to resonance by placing the magnetic field along the [100] crystalline axis. 
%Other example of class resonances are given in the SI.
This configuration will be essential to decipher what dipolar relaxation mechanisms play a role at low field.

Fig. \ref{100_VS_1x4} (b)-ii) and (b)-iii) show the PL from an ensemble of NV centers as a function of the amplitude of a magnetic field aligned as in Fig. \ref{100_VS_1x4} (b)-i). 
We also extract the stretched exponential spin decay time $T_1^{\rm dd}$ by running the sequence shown in Fig. 2-c) at each of the XX magnetic field values. 
We observe that the 3 \% increase in PL under low magnetic field is indeed correlated with a drop in the spin decay rate from 1.6 kHz to 0.05 kHz, as the magnetic field increases. This result can be explained by the higher number of resonant spins in zero magnetic field, where all four classes are degenerate, as opposed when $\bm B \neq 0$. The half width of the dip ($\approx 10$G) is consistent with a fluctuator model taking into account the change in the spin resonance overlaps as the $B$ field is swept (see SI ?).
Note that the decrease in PL for $B>40\ \rm G$ are related to state mixing by the transverse magnetic field, so it is not correlated to a modification of the spin lifetime.

Importantly, we observed that the lifted degeneracy of the four NV classes is not the sole reason for the increase of the spin lifetime as the B field increases. Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii) show the change in the PL and in $1/T_1^{\rm dd}$ as a function of a magnetic field that is aligned along the [100] axis. For this particular orientation, the four classes are always resonant, regardless of the field amplitude. We can see that there still is an increase in the spin decay rate and a corresponding drop in the PL under low magnetic field values, although considerably reduced. Also, the rate of change in both the PL and $T_1^{\rm dd}$ is larger by more than a factor of 3 in the [100] configuration. Although the contrast is smaller, this effect can thus drastically impact zero-field magnetometry protocols based on CR.
The main aim of this paper is to understand the physics governing the contribution to the spin decay in this second scenario. Note that the slight drop in PL and the corresponding bump for $1/T_1^{\rm dd}$ at $B \sim 20$G is related to dipolar interaction with NV centers that have a $^{13}$C as a first neighbor \cite{pellet2021optical}.

We identified and isolated two additional mechanisms that could explain this effect : local electric fields and double-flip processes. 
The dipole-dipole interaction Hamiltonian between two spins ${\vec S}_1$ and ${\vec S}_2$ reads :
\begin{equation}
\mathcal{H}^{\rm dd}= -\frac{J_0}{r^3}\left(3\left({\vec S}_1 \cdot \vec u \right)\left({\vec S}_2 \cdot \vec u \right) - {\vec S}_1 \cdot {\vec S}_2  \right),
\end{equation}
Where $J_0= (2 \pi) 52\ \rm MHz \cdot \rm{nm}^3$, $\vec r$ is the relative positions of the two spins and $\vec u = \frac{\vec r}{\norm{\vec r}}$. In situations such as the one described in Fig.XX, the only relevant terms of $\mathcal{H}^{\rm dd}$ in terms of population transfer are the flip-flop terms such as $\mel{0,+1}{\mathcal{H}^{\rm dd}}{+1,0}$. In zero magnetic field however, we have to take into account these two other mechanisms of the dipolar Hamiltonian in consideration.
%where both NV spins involved in the dipole-dipole coupling flip their spin in the same direction.
It was shown in  \cite{mittiga2018imaging} that local electric fields coming from charge traps such as $P_1^+$ centers are responsible for the optically detected spin resonance profile at zero magnetic field in large NV ensembles. Magnetic field noise comes in to second order in zero field. %We also do not take into account the hyper-fine structure of the NV center due to the large inhomogeneous broadening of the ESR transitions (see SI). 
We will then consider the following spin Hamiltonian for the electric field dependent NV$^-$ ground state : 
\begin{equation}
\label{NV Hamiltonian}
\mathcal{H}_{\rm elec}= d_\perp \left[ E_x(S_y^2-S_x^2) + E_y(S_xS_y+S_yS_x) \right]
\end{equation}
where $E_{x,y}$ are the projections of local electric fields on the NV axes. 
Under local electric fields with orientations given by the angle $\phi_E={\rm tan}(E_x/E_y)$, the eigenstates of $H_{NV}$ are $\ket{0}$ and $\ket{\pm}=\frac{1}{\sqrt{2}}(\ket{+1}\pm e^{-i\phi_E}\ket{-1})$.
Computing the flip-flop terms $\ket{\pm,0} \bra{0,\pm} $ in the fluctuator model shows that $1/T_1^{dd}$ should be increased at zero field, even after averaging over all possible orientations $\phi_E$ (see SI).  Another mechanism at stake when considering dipolar interactions with spin-1 systems are double-flip processes. 
These processes correspond to terms like $\ket{+1,0} \bra{0,1}$ in the anisotropic dipolar interaction, giving rise to an exchange of two units of spin-angular momentum. These processes can become significant when the states $\pm 1$ are degenerate and are thus likely to explain part of the zero-field dip in Fig. c)-ii). 
%Theoretical model predicts a factor of $\approx $5 increase in $1/T_1^{dd}$ at zero field.

In order to discriminate the role of these two effects in the low-field region, we performed another similar study under a purely transverse magnetic field.
This regime in fact emulates the effect of the electric field while turning off the resonance condition for double flip processes : indeed for relatively weak transverse magnetic field, the eigen-basis of the spin Hamiltonian is close to $\ket{0},\ket{+},\ket{-}$. This property has been used to observe a linear dependence in the electric field for the spin transition frequencies by applying an external transverse magnetic field \cite{dolde2011electric,qiu2022nanoscale}.
Fig.~\ref{B_transverse} (a) shows the energy levels for the three eigenstates of the NV spin Hamiltonian as a function of a purely transverse magnetic field. We will denote these eigenstates in the general case $\ket{g}$, $\ket{d}$ and $\ket{e}$. For weak transverse magnetic field, $\ket{g} \approx \ket{0}$, $\ket{d}=\ket{-}$ and $\ket{e} \approx \ket{+}$. As the magnetic field increases, $\ket{g}$ and $\ket{e}$ starts to become a mixing of $\ket{0}$ and $\ket{+}$ while $\ket{d}$ remains equal to $\ket{-}$.

Crucially, mixing remains even when the associated energy splitting of the $\ket{d}$ and $\ket{e}$ states is large : Fig. \ref{B_transverse}-(b) shows both the splitting between the energy levels of $\ket{d}$ and $\ket{e}$, and the matching factor $\abs{\bra{e}\ket{+}}^2$ characterizing the closeness of the states $\bra{e}$ and $\bra{+}$. We can see that at a field value of $\sim 130$ G, the splitting between the $\ket{d}$ and $\ket{e}$ reaches $\sim 50$ MHz, far exceeding the range of the dipole-dipole resonant coupling (see details in SI), but $\abs{\bra{e}\ket{+}}^2 > 0.98$ meaning that the spin Hamiltonian eigenstates are still roughly in the $\ket{0}, \ket{+}, \ket{-}$ basis. We can therefore decouple the effects of the double-flip processes  $\propto \ket{g,d}\bra{d,e}$, which are no longer resonant when $| \bm B | > 130$ G, from the effects of the mixing induced by the transverse electric - or in this case magnetic - field.

Fig. \ref{B_transverse}-(c) shows the measurement of the stretched lifetime $T_1^{\rm dd}$ for a single class of NV centers exposed to a transverse magnetic field between 25 (??) and 130 G (FOR A DIFFERENT SAMPLE... which ? we observe a $1/T_1^{\rm dd}$ of $30$ Hz under large longitudinal magnetic fields for this class ?). The detuning $\Delta \nu$ between the two states $\ket{d}$ and $\ket{e}$ was measured through ODMR. We attribute the decrease in the spin decay rate when the detuning increases to the loss of effectiveness of the double-flip processes. We can notice that the decay rate reaches a plateau for $\Delta \nu \geq 30\ \rm MHz$ with a value about twice as large as the decay rate for a longitudinal magnetic field. This plateau we attribute to the effect of the transverse magnetic field, which we assume to be similar to the effect of the electric field in the low magnetic field regime. We should note that the effect of the double-flip processes is about 4 times more important in the depolarization rate than the effect of the electric field, even though the states are not initially fully resonant.

Based on an extension of the fluctuator model developed in SI, we predict an overall increase by a factor of $\sim 4$ of the dipole-dipole induced decay rate in the non-magnetic basis $\ket{0}, \ket{+}, \ket{-}$ compared to the magnetic basis $\ket{0}, \ket{+1}, \ket{-1}$. While the measured increase is smaller than the predicted one, the measurements agrees with the theoretical increase in the flip-flop rate in the $\ket{+/-}$ basis compared to the $\ket{\pm 1}$ basis.

\begin{figure}
\includegraphics[width=0.45\textwidth]{Figures/fig_transverse_field_V2.pdf}
\caption{(a) Energies for the spin Hamiltonian in the presence of pure transverse magnetic field $B_\perp$. The three spin levels are denoted $\ket{g}$, $\ket{d}$ and $\ket{e}$. (b) Blue dashed curve : frequency detuning $\Delta \nu$ between the $\ket{e}$ and $\ket{d}$ states. Red plain curve : Matching factor $\abs{\bra{e}\ket{+}}^2$ between the states $\bra{e}$ and $\bra{+}$. (c) Measurement of the stretched exponential decay rate $1/T_1^{\rm dd}$ for a single class as a function of the measured detuning $\Delta \nu$ between the two transitions $\ket{g} \to \ket {d}$ and  $\ket{g} \to \ket {e}$ for purely transverse magnetic field. The green dashed line correspond to the decay rate for the same class when the magnetic field is longitudinal}
\label{B_transverse}
\end{figure}
Similarly, we compute the predicted decay rate when $\bm B \parallel \left[100\right]$ case compared when $\bm B = 0$, and find a theoretical increase of $\sim 20\%$ when $\bm B = 0$ due to the mixing by the local electric field. This, along with the double-flip processes explains the behavior observed in Fig. \ref{100_VS_1x4} (c)-ii) and (c)-iii).
